---
title: "Introduction to DIAlignR"
author: "Shubham Gupta, Hannes Rost"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this document we are presenting workflow of DIAlignR for retention time (RT) alignment of targeted proteomics (e.g. DIA, SWATH-MS, PRM, SRM) runs. This tool works with MS2 chromatograms and provides a hybrid approach of global and local alignment to establish correspondence between peaks.

# Install DIAlignR
```{r installDIAlignR, eval=FALSE}
require(devtools)
install_github("Roestlab/DIAlignR")
```

```{r loadDIAlignR}
library(DIAlignR)
```

# Citation
Gupta S, Ahadi S, Zhou W, Röst H. "DIAlignR Provides Precise Retention Time Alignment Across Distant Runs in DIA and Targeted Proteomics." Mol Cell Proteomics. 2019 Apr;18(4):806-817. doi: https://doi.org/10.1074/mcp.TIR118.001132 

# Prepare chromatograms from Spectra
Mass-spec output file mostly contains spectra. Targeted proteomics works on identifying analytes from their chromatographic elution profile. DIAlignR extends the same concept for retention-time (RT) alignment. For RT alignment DIAlignR relies on MS2 chromatograms. Open-source software tools e.g. OpenSWATH can be used to extract features and extracted-ion chromatograms (XICs) from spectra file. A detailed [tutorial](http://openswath.org/en/latest/docs/openswath.html) to run OpenSWATH is availabe. In short, following command can be used:
```{bash, eval=FALSE}
./bin/OpenSwathWorkflow -in Spectra.mzML.gz -tr library.TraML -
swath_windows_file swathsWindows.txt -sort_swath_maps -tr_irt
iRTassays.TraML -out_tsv Feature.tsv -out_chrom XIC.chrom.mzML -readOptions cacheWorkingInMemory -tempDirectory temp/ -min_upper_edge_dist 1
```

Output files **Feature.tsv** and **XIC.chrom.mzML** are needed to perform hybrid alignment through DIAlignR. Some chromatograms are stored in compressed form and could be difficult to have accessed by `mzR`. In such cases `mzR` would throw an error indicating `Invalid cvParam accession "1002746"`. To avoid this issue, uncompress chromatograms using OpenMS.
```{bash, eval=FALSE}
./bin/FileConverter -in XIC.chrom.mzML -in_type 'mzML' -out XIC.chrom.mzML
```

# Load MS2 chromatograms using mzR
MS2 chromatograms can be extracted from mzML files using [mzR](https://bioconductor.org/packages/release/bioc/html/mzR.html) package. The workflow of extraction of a few chromatograms for this paper is provided below. From the FTP link provided [here](http://www.peptideatlas.org/PASS/PASS01280) download `mzML` files from these directories:   

* Spyogenes/raw/hroest_K120808_Strep10PlasmaBiolRepl2_R02_SW   
* Spyogenes/raw/hroest_K120808_Strep0PlasmaBiolRepl2_R01_SW    

Few precursor IDs and their transition IDs are tabulated in this file at the same FTP link:

* Spyogenes/assay/500Peptide4Alignment.csv      
```{r loadChroms, eval=FALSE}
library(mzR)
library(signal)
TargetPeptides <- read.table("500Peptide4Alignment.csv", sep = ",", header = T)
temp <- list.files(pattern="*.mzML", recursive = TRUE)
for(filename in temp){
  # This makes sure that order of extracted MS2 chromatograms is same for each run.
  mz <- openMSfile(filename, backend = "pwiz")
  chromHead <- chromatogramHeader(mz)
  filename <- gsub("(.*)(/hroest_)(.*)(_SW.chrom.mzML)", replacement = "\\3", filename)
  chromatogramIndices <- chromHead$chromatogramIndex[match(TargetPeptides$transition_name, chromHead$chromatogramId)]
  TargetPeptides[filename] <- chromatogramIndices
  transition_group_ids <- unique(TargetPeptides$transition_group_id)
  ChromsExtractedPerRun <- sapply(transition_group_ids, function(id){
    chromIndices <- chromatogramIndices[TargetPeptides$transition_group_id == id]
    # ChromsExtracted <- lapply(1:length(chromIndices), function(i) chromatograms(mz, chromIndices[i]))
    ChromsExtracted <- lapply(1:length(chromIndices), function(i) {
      rawChrom <- chromatograms(mz, chromIndices[i])
      rawChrom[,2] <- sgolayfilt(rawChrom[,2], p = 4, n = 9) # To smooth chromatograms, use Savitzky-Golay filter
      return(rawChrom)
    } )
    return(ChromsExtracted)
  })
  names(ChromsExtractedPerRun) <- transition_group_ids
  rm(mz)
  saveRDS(ChromsExtractedPerRun, paste0(filename, "_ChromSelected.rds"))
}
write.table(TargetPeptides, file = "TargetPeptidesWchromIndex.csv", sep = ",")

# Load chromatograms of all runs 
temp <- list.files(pattern = "*_ChromSelected.rds")
StrepChroms1 <- list()
for(i in 1:length(temp)){
  StrepChroms1[[i]] <- readRDS(temp[i])
}
temp <- sapply(temp, strsplit, split = "_ChromSelected.rds", USE.NAMES = FALSE)
names(StrepChroms1) <- temp
```

# Dummy data

We have provided MS2 chromatograms of 20 precursors across four runs with this package.     
1. `StrepChroms` contains MS2 chromatograms for precursors.     
2. `runnames` has names of runs matched with data in  [this repo](http://www.peptideatlas.org/PASS/PASS01280).     
3. `StrepAnnot` has manual annotation of precursor peaks.     
4. `oswOutStrep` has picked featured from each run using **OpenSWATH**.    
5. `peptides` has names of all 20 precursors.
```{r, eval=FALSE}
runs <- names(StrepChroms);
peptides <- names(StrepChroms[["run1"]])
```

## How do unaligned peak-groups look like?

Here we present XICs of precursor `3505_LATWYSEMK/2` in `run3` and `run4`. In MS/MS, each precursor has a number of fragment traces. These traces are shown by different colors. co-elution of all fragment traces is annotated as true-peak. The peak is identified at t = 3513.6 sec in `run3` and and at t = 3606.0 sec `run4`.

## Visualize chromatograms, similarity matrix and alignment path
```{r, echo = FALSE, warning=FALSE}
library(lattice)
library(ggplot2)
library(reshape2)
library(zoo)

plotChromatogram <- function(data, run, peptide, peakAnnot, printTitle =TRUE){
  df <- do.call("cbind", data[[run]][[peptide]])
  df <- df[,!duplicated(colnames(df))]
  df <- melt(df, id.vars="time", value.name = "Intensity")
  g <- ggplot(df, aes(time, Intensity, col=variable)) + geom_line(show.legend = FALSE) + theme_bw()
  if(printTitle) g <- g + ggtitle(paste0(run, ", ",peptide)) + theme(plot.title = element_text(hjust = 0.5))
  g <- g + geom_vline(xintercept=peakAnnot[peptide, run], lty="dotted", size = 0.4)
  return(g)
}
plotSingleAlignedChrom <- function(data, run, peptide, idx, t, printTitle = TRUE){
  intensity <- list()
  for(k in 1:length(data[[run]][[peptide]])){
      mutateInt <- data[[run]][[peptide]][[k]][idx, 2]
      mutateInt <- na.locf(na.locf(mutateInt, na.rm = FALSE),fromLast = TRUE)
      intensity[[k]] <- mutateInt
  }
  df <- do.call("cbind", intensity)
  Index <- 1:nrow(df)
  df <- cbind(Index, as.data.frame(df))
  df <- melt(df, id.vars="Index", value.name = "Intensity")

  g <- ggplot(df, aes(Index, Intensity, col=variable)) + geom_line(show.legend = FALSE) + theme_bw()
  if(printTitle) g <- g + ggtitle(paste0(run, ", ",peptide)) + theme(plot.title = element_text(hjust = 0.5))
  return(g)}
```


```{r VisualizeAlignment, fig.width=6, fig.align='center', fig.height=4, eval=TRUE, fig.show = "hold"}
plotChromatogram(StrepChroms, "run3", "3505_LATWYSEMK/2", StrepAnnot)
plotChromatogram(StrepChroms, "run4", "3505_LATWYSEMK/2", StrepAnnot)
```

# Establish correspondence between peaks
Although, peaks in XICs can identified manually, it is unscalable and biased. We aim to have consistent peak picking by establishing pairwise correspondence. Pairwise correspondence is achieved by aligning retention-time of two runs. Following approaches are discussed in the literature:

## Fit a global alignment function between runs
A simple approach is to fit a liner or non-linear function that takes into account all peaks detected by peak-picking software. On our dummy dataset, we establish correspondence by having non-linear LOESS alignment. `getLOESSfit` and `getPepPeakCorp` functions from the package can be used to fit LOESS. It takes in OpenSWATH output features `oswOutStrep` and returns a non-linear fit. For testing test `peptides` are given so that `getLOESSfit` excludes them from training data. The average error for LOESS alignment is 4.051 sec.

```{r globalFit, eval=TRUE}
run_pair <- "run1_run2"
loess.fit <- getLOESSfit(run_pair, peptides, oswOutStrep, 0.15)
StrepAnnot <- as.data.frame(StrepAnnot)
predict.run2 <- predict(loess.fit, data.frame(RUN1 = StrepAnnot[, "run1"]))
Err_global <- predict.run2 - StrepAnnot[,"run2"]
sum(abs(Err_global))/20
MappedTimeGlobal <- getPepPeakCorp(featureTable =  StrepAnnot, pairName = run_pair, alignType = "global", oswFeatureList = oswOutStrep, spanvalue = 0.15)
Err_global <- StrepAnnot[,"run2"] - MappedTimeGlobal
sum(abs(Err_global))/20
```

## Fit a local alignment between chromatogram groups
Another approach is to align each precursor without the influence of remaining precursors. This is termed as **local alignment**. `getPepPeakCorp` function aligns chromatogram-groups of a precursor non-linearly using dynamic programming. For 20 test peptides, the local alignment has average error of 2.46 sec. 
```{r localFit, eval=TRUE}
simMeasure <- "dotProductMasked"
pair <- "run1_run2"
MappedTimeLocal <- getPepPeakCorp(StrepAnnot, pair, StrepChroms[["run1"]], StrepChroms[["run2"]], "local", oswOutStrep)
Err_local <- StrepAnnot[,"run2"] - MappedTimeLocal
sum(abs(Err_local))/20
```

## Hybrid alignment of chromatograms
We also provide a hybrid (local+global) approach which provides accuracy of local alignment and robustness of global alignment. `getPepPeakCorp` function uses global alignment to limit search space of `local` alignment. The average error for hybrid alignment is 2.46 sec.
```{r hybridAlignParam}
samplingTime <-3.4 # In example dataset, all points are acquired at 3.4 second interval.
meanRSE <- 7.4 # Defines base upper limit of sampling space within that local alignment is done. 
pair <- "run1_run2"
MappedTimeHybrid <- getPepPeakCorp(StrepAnnot, pair, StrepChroms[["run1"]], StrepChroms[["run2"]], "hybrid", oswOutStrep, expRSE = meanRSE, samplingTime = samplingTime)
Err_hybrid <- StrepAnnot[,"run2"] - MappedTimeHybrid
sum(abs(Err_hybrid))/20
```

## plot fraction of aligned peptides
From the figure below, we can see that `hybrid alignment` approach aligns more peptides within alignment error of 10 sec. 
```{r}
plotErrorCurve <- function(x, clr = "black", SameGraph = FALSE, xmax = 120, ...){
    x <- x[!is.na(x)]
    breaks = seq(0, xmax, by=0.5)
    duration.cut = cut(x, breaks, right = FALSE) 
    duration.freq = table(duration.cut)
    cumfreq0 = c(0, cumsum(duration.freq))
    if(SameGraph == TRUE){lines(breaks, cumfreq0/length(x), col = clr, ...)}
    else{plot(breaks, cumfreq0/length(x), col = clr, type = "l", ...)}
}
```

```{r plotErr, fig.width=6, fig.align='center', fig.height=6, fig.show='hold', eval=TRUE}
plotErrorCurve(abs(Err_global), "blue",xmax = 60, xlab = "Retention time difference (in sec)", ylab = "Cumulative fraction of peptides")
plotErrorCurve(abs(Err_local), "darkgreen", SameGraph = TRUE, xlab = "Retention time difference (in sec)", ylab = "Cumulative fraction of peptides")
plotErrorCurve(abs(Err_hybrid), "red", SameGraph = TRUE, xlab = "Retention time difference (in sec)", ylab = "Cumulative fraction of peptides")
grid(); legend("bottomright",  c("global", "local", "hybrid"), col = c("blue", "darkgreen", "red"), lwd = 1.5, box.lwd = 1)
```


# How do these alignment approaches differ?
## Global alignment
Global alignment approach fits a global function using all detected peaks. In the figure below, blue line represents alignment function. Two test petides (solid green dots) have their retention times mapped from Run1 to Run2 using this function.     
The two precursors are `15605_YFMPVHGEYR/3` and `9091_NDYGNTTLALR/2`.
```{r, fig.width=6, fig.align='center', fig.height=6, echo=FALSE}
run_pair <- c("run1", "run2")
RUN1 <- oswOutStrep[[run_pair[1]]]; RUN2 <- oswOutStrep[[run_pair[2]]]
cmp <- intersect(RUN1[,1], RUN2[,1]) # First column corresponds to transition_group_record
RUN1 <- RUN1[which(RUN1[,1] %in% cmp), ]
RUN2 <- RUN2[which(RUN2[,1] %in% cmp), ]
RUN1 <- RUN1[match(cmp, RUN1[,1]),]
RUN2 <- RUN2[match(cmp, RUN2[,1]),]
RUNS_RT <- data.frame( "transition_group_record" = RUN1[,1], "RUN1" = RUN1$RT, "RUN2" = RUN2$RT)
RUNS_RT <- RUNS_RT[order(RUNS_RT$RUN1), ]
testPeptides <-intersect(cmp, peptides)
plot(RUNS_RT[,"RUN1"], RUNS_RT[,"RUN2"], xlim = c(2820,2980), ylim = c(2820,2980), xlab = "Run1 retention time (sec)", ylab = "Run2 retention time (sec)", pch=4, cex = 0.4, main = "Non-linear global alignment")
# For testing we want to avoid validation peptides getting used in the fit.
Loess.fit <- loess(RUN2 ~ RUN1, data = RUNS_RT, subset = !transition_group_record %in% testPeptides,
                   span = 0.15, control=loess.control(surface="direct"))
lines(RUNS_RT[,"RUN1"], predict(Loess.fit, newdata=RUNS_RT[,"RUN1"]), col = "blue")
points(StrepAnnot$run1[c(18,19)], StrepAnnot$run2[c(18,19)], col = "red", pch = 19)
lines(RUNS_RT[,"RUN1"], predict(Loess.fit, newdata=RUNS_RT[,"RUN1"])-2*Loess.fit$s, lty = "dashed")
lines(RUNS_RT[,"RUN1"], predict(Loess.fit, newdata=RUNS_RT[,"RUN1"])+2*Loess.fit$s, lty = "dashed")
grid(); legend("topleft", c("LOESS fit", "2*RSE"), lty= c("solid", "dashed"), col = c("blue", "black"))
```

## Local alignment
Local alignment uses raw MS2 chromatograms for alignment. Unaligned XICs of precursors "15605_YFMPVHGEYR/3" are shown below.
```{r, fig.width=6, fig.align='center', fig.height=4, fig.show='hold'}
plotChromatogram(StrepChroms, "run1", "15605_YFMPVHGEYR/3", StrepAnnot)
plotChromatogram(StrepChroms, "run2", "15605_YFMPVHGEYR/3", StrepAnnot)
```

```{r, echo=FALSE}
peptide <- "15605_YFMPVHGEYR/3"
simMeasure <- "dotProductMasked"
intensityListA <- lapply(StrepChroms[["run1"]][[peptide]], `[[`, 2) # Extracting intensity values
intensityListB <- lapply(StrepChroms[["run2"]][[peptide]], `[[`, 2) # Extracting intensity values
tA <- StrepChroms[["run1"]][[peptide]][[1]][["time"]] # Extracting time component
tB <- StrepChroms[["run2"]][[peptide]][[1]][["time"]] # Extracting time component
localObj <- alignChromatogramsCpp(intensityListA, intensityListB, "local", tA, tB, "mean", simMeasure)
```

At first, we build a similarity matrix for these XICs. Then we find alignment path through this similarity matrix using dynamic programming.
```{r localAlign, fig.width=6, fig.align='center', fig.height=4}
s <- localObj@s
Path <- localObj@path[2:nrow(localObj@path), 2:ncol(localObj@path)]
levelplot(s, axes = TRUE, xlab = "run1 index", ylab = "run2 index")
levelplot(s, axes = TRUE, xlab = "run1 index", ylab = "run2 index", main = paste0("Local alignment through the similarity matrix\n for ", peptide), fontsize = 7) + latticeExtra::as.layer(levelplot(Path, col.regions = c("transparent", "green"), alpha = 1, axes = FALSE))
```


Following the path, we have aligned time-points through which we can develop aligned chromatograms.
```{r, fig.width=6, fig.align='center', fig.height=4, fig.show='hold'}
pTR <- plotSingleAlignedChrom(StrepChroms, "run1", peptide, localObj@indexA_aligned, tA, TRUE) + geom_vline(xintercept=which.min(abs(tA - StrepAnnot[peptide, "run1"])), lty="dotted", size = 0.4)
pBR <- plotSingleAlignedChrom(StrepChroms, "run2", peptide, localObj@indexB_aligned, tB, TRUE) + geom_vline(xintercept=which.min(abs(tB - StrepAnnot[peptide, "run2"])), lty="dotted", size = 0.4)
pBR <- pBR + geom_vline(xintercept=which.min(abs(tA - StrepAnnot[peptide, "run1"])), lty="dashed", size = 0.4, color = "red")
pTR
pBR
```


## Hybrid alignment
In hybrid alignment, we follow similar steps like local alignment, except for, before finding path we penalize similarity matrix.
```{r hybridAlign, fig.width=6, fig.align='center', fig.height=4}
hybridObj <- getAlignedObj(peptide, pair, StrepChroms[["run1"]], StrepChroms[["run2"]], "hybrid", oswOutStrep, samplingTime = 3.4)
s <- hybridObj@s
Path <- hybridObj@path[2:nrow(hybridObj@path), 2:ncol(hybridObj@path)]
levelplot(s, axes = TRUE, xlab = "run1 index", ylab = "run2 index")
levelplot(s, axes = TRUE, xlab = "run1 index", ylab = "run2 index", main = paste0("Hybrid alignment through the similarity matrix\n for ", peptide), fontsize = 7) + latticeExtra::as.layer(levelplot(Path, col.regions = c("transparent", "green"), alpha = 1, axes = FALSE))
```


# Citation
Gupta S, Ahadi S, Zhou W, Röst H. "DIAlignR Provides Precise Retention Time Alignment Across Distant Runs in DIA and Targeted Proteomics." Mol Cell Proteomics. 2019 Apr;18(4):806-817. doi: https://doi.org/10.1074/mcp.TIR118.001132 

# Session Info
```{r sessionInfo, eval=TRUE}
devtools::session_info()
```

<br>

## Last compilation
Last compiled at `r Sys.Date()`.
