---
title: "Introduction to DIAlign package for alignment of targeted proteomics runs"
author: "Shubham Gupta, Hannes Rost"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this document we are presenting workflow of DIAlign for retention time (RT) alignment of targeted proteomics (e.g. DIA, SWATH-MS) runs. This tool works with MS2 chromatograms and uses a hybrid approach of global and local alignment to establish correspondence between peaks.

# Install DIAlign 
```{r installDIAlign}
require(devtools)
install_github("Roestlab/DIAlign")
library(DIAlign)
```


# Fit a global alignment function between runs
```{r globalFit}
run_pair <- c("run1", "run2")
loess.fit <- getLOESSfit(run_pair, peptides, oswOutStrep, 0.15)
StrepAnnot <- as.data.frame(StrepAnnot)
predict.run2 <- predict(loess.fit, data.frame(RUN1 = StrepAnnot[, run_pair[1]]))
Err <- predict.run2 - StrepAnnot[,run_pair[2]]
```


## plot fraction of aligned peptides
```{r plotGlobal, fig.width=6, fig.align='center', fig.height=6}
plotErrorCurve <- function(x, clr = "black", SameGraph = FALSE, xmax = 120, ...){
    x <- x[!is.na(x)]
    breaks = seq(0, xmax, by=0.5)
    duration.cut = cut(x, breaks, right = FALSE) 
    duration.freq = table(duration.cut)
    cumfreq0 = c(0, cumsum(duration.freq))
    if(SameGraph == TRUE){lines(breaks, cumfreq0/length(x), col = clr, ...)}
    else{plot(breaks, cumfreq0/length(x), col = clr, type = "l", ...)}
}
plotErrorCurve(abs(Err), "blue", xlab = "Retention time difference (in sec)", ylab = "Cumulative fraction of peptides")
```


# Fit a local alignment between chromatogram groups
```{r localFit, eval=TRUE}
gapQuantile <- 0.5; goFactor <- 1/8; geFactor <- 40
simMeasure <- "dotProductMasked"
run_pair <- c("run1", "run2")
Err <- matrix(NA, nrow = length(peptides), ncol = 1)
rownames(Err) <- peptides
for(peptide in peptides){
  s <- getSimilarityMatrix(StrepChroms, peptide, run_pair[1], run_pair[2], type = simMeasure)
  gapPenalty <- getGapPenalty(s, gapQuantile, type = simMeasure)
  Alignobj <- getAffineAlignObj(s, go = gapPenalty*goFactor, ge = gapPenalty*geFactor)
  AlignedIndices <- getAlignment(Alignobj)
  tA <- StrepChroms[[run_pair[1]]][[peptide]][[1]][["time"]]
  tB <- StrepChroms[[run_pair[2]]][[peptide]][[1]][["time"]]
  tA.aligned <- mapIdxToTime(tA, AlignedIndices[[1]][,"indexA_aligned"])
  tB.aligned <- mapIdxToTime(tB, AlignedIndices[[1]][,"indexB_aligned"])
  predictTime <- tB.aligned[which.min(abs(tA.aligned - StrepAnnot[peptide, run_pair[1]]))]
  deltaT <- predictTime - StrepAnnot[peptide, run_pair[2]]
  Err[peptide, 1] <- deltaT
}
```


```{r plotLocal, fig.width=6, fig.align='center', fig.height=6, fig.show='hold'}
plotErrorCurve(abs(Err), "darkgreen", SameGraph = FALSE, xlab = "Retention time difference (in sec)", ylab = "Cumulative fraction of peptides")
```


# Hybrid alignment of chromatograms
```{r hybridAlignParam}
samplingTime <-3.4 # In example dataset, all points are acquired at 3.4 second interval.
samples4gradient <- 100; RSEdistFactor <- 3.5; hardConstrain <- FALSE
pair_names <- vector(); runs <- names(StrepChroms)
for (i in 1:(length(runs)-1)){
    for (j in (i+1): length(runs)){
        pair_names <- c(paste(runs[i], runs[j], sep = "_"), pair_names)
    }}
globalStrep <- matrix(NA, nrow = 1, ncol = length(pair_names))
colnames(globalStrep) <- pair_names
rownames(globalStrep) <- c("RSE")

for(pair in pair_names){
  run_pair <- strsplit(pair, split = "_")[[1]]
  Loess.fit <- getLOESSfit(run_pair, peptides, oswOutStrep, 0.1)
  globalStrep["RSE", pair] <- Loess.fit$s
}
meanRSE <- mean(globalStrep["RSE",])
```


```{r hybridAlign, eval=TRUE}
gapQuantile <- 0.5; goFactor <- 1/8; geFactor <- 40
simMeasure <- "dotProductMasked"
run_pair <- c("run1", "run2"); pair <- "run1_run2"
Err <- matrix(NA, nrow = length(peptides), ncol = 1)
rownames(Err) <- peptides
Loess.fit <- getLOESSfit(run_pair, peptides, oswOutStrep, 0.1)
for(peptide in peptides){
  s <- getSimilarityMatrix(StrepChroms, peptide, run_pair[1], run_pair[2], type = simMeasure)
  gapPenalty <- getGapPenalty(s, gapQuantile, type = simMeasure)
  tRunAVec <- StrepChroms[[run_pair[1]]][[peptide]][[1]][["time"]]
  tRunBVec <- StrepChroms[[run_pair[2]]][[peptide]][[1]][["time"]]
  noBeef <- ceiling(RSEdistFactor*min(globalStrep["RSE", pair], meanRSE)/samplingTime)
  if(hardConstrain) {
    MASK <- calcNoBeefMaskGlobal(tRunAVec, tRunBVec, Fit = Loess.fit, noBeef)
    s <- constrainSimilarity(s, MASK, -2*max(s))
    } else {
      MASK <- calcNoBeefMaskGlobalWSlope(tRunAVec, tRunBVec, Fit = Loess.fit, noBeef)
      s <- constrainSimilarity(s, MASK, -2*max(s)/samples4gradient) # it will take 100 time points to reach -2max
    }
  Alignobj <- getAffineAlignObj(s, go = gapPenalty*goFactor, ge = gapPenalty*geFactor)
  AlignedIndices <- getAlignment(Alignobj)
  tA <- StrepChroms[[run_pair[1]]][[peptide]][[1]][["time"]]
  tB <- StrepChroms[[run_pair[2]]][[peptide]][[1]][["time"]]
  tA.aligned <- mapIdxToTime(tA, AlignedIndices[[1]][,"indexA_aligned"])
  tB.aligned <- mapIdxToTime(tB, AlignedIndices[[1]][,"indexB_aligned"])
  predictTime <- tB.aligned[which.min(abs(tA.aligned - StrepAnnot[peptide, run_pair[1]]))]
  deltaT <- predictTime - StrepAnnot[peptide, run_pair[2]]
  Err[peptide, 1] <- deltaT
}
```

```{r plotHybrid, fig.width=6, fig.align='center', fig.height=6}
plotErrorCurve(abs(Err), "red", SameGraph = FALSE, xlab = "Retention time difference (in sec)", ylab = "Cumulative fraction of peptides")
```

## Visualize chromatograms, similarity matrix and alignment path
```{r VisualizeAlignment, fig.width=6, fig.align='center', fig.height=6}
library(lattice)
library(ggplot2)
library(reshape2)

plotChromatogram <- function(data, run, peptide, StrepAnnot, printTitle =TRUE){
  df <- do.call("cbind", data[[run]][[peptide]])
  df <- df[,!duplicated(colnames(df))]
  df <- melt(df, id.vars="time", value.name = "Intensity")
  g <- ggplot(df, aes(time, Intensity, col=variable)) + geom_line(show.legend = FALSE) + theme_bw()
  if(printTitle) g <- g + ggtitle(paste0(run, ", ",peptide)) + theme(plot.title = element_text(hjust = 0.5))
  g <- g + geom_vline(xintercept=StrepAnnot[peptide, run], lty="dotted", size = 0.4)
  return(g)
}

levelplot(s, axes = TRUE, xlab = "run1 index", ylab = "run2 index")
Path <- getAlignmentPath(AlignedIndices[[1]], s)
levelplot(s, axes = TRUE, xlab = "run1 index", ylab = "run2 index", main = paste0("Alignment path through the ", simMeasure, " similarity matrix\n for ", peptide)) + latticeExtra::as.layer(levelplot(Path, col.regions = c("transparent", "green"), alpha = 1, axes = FALSE))
```


# Session Info
```{r sessionInfo, eval=TRUE}
devtools::session_info()
```

<br>

## Last compilation
Last compiled at `r Sys.Date()`.
