---
title: "Introduction to DIAlign package for alignment of targeted proteomics runs"
author: "Shubham Gupta, Hannes Rost"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this document we are presenting workflow of DIAlign for retention time (RT) alignment of targeted proteomics (e.g. DIA, SWATH-MS) runs. This tool work with MS2 chromatograms and uses a hybrid approach of global and local alignment to extablish correspondence between signals.

# Install DIAlign 
```{r installDIAlign}
require(devtools)
install_github("Roestlab/DIAlign")
library(DIAlign)
```


# Fit a global alignment function between runs
```{r globalFit}
run_pair <- c("run1", "run2")
loess.fit <- getLOESSfit(run_pair, peptides, oswOutStrep, 0.15)
StrepAnnot <- as.data.frame(StrepAnnot)
predict.run2 <- predict(loess.fit, data.frame(RUN1 = StrepAnnot[, run_pair[1]]))
Err <- predict.run2 - StrepAnnot[,run_pair[2]]
```


## plot fraction of aligned peptides
```{r plotGlobal}
plotErrorCurve <- function(x, clr = "black", SameGraph = FALSE, xmax = 120, ...){
    x <- x[!is.na(x)]
    breaks = seq(0, xmax, by=0.5)
    duration.cut = cut(x, breaks, right = FALSE) 
    duration.freq = table(duration.cut)
    cumfreq0 = c(0, cumsum(duration.freq))
    if(SameGraph == TRUE){lines(breaks, cumfreq0/length(x), col = clr, ...)}
    else{plot(breaks, cumfreq0/length(x), col = clr, type = "l", ...)}
}
plotErrorCurve(abs(Err), "blue", xlab = "Retention time difference (in sec)", ylab = "Cumulative fraction of peptides")
```


# Fit a local alignment between chromatogram groups
```{r localFit}
gapQuantile <- 0.5; goFactor <- 1/8; geFactor <- 40
simMeasure <- "dotProductMasked"
run_pair <- c("run1", "run2")
Err <- matrix(NA, nrow = length(peptides), ncol = 1)
rownames(Err) <- peptides
for(peptide in peptides){
  s <- getSimilarityMatrix(StrepChroms, peptide, run_pair[1], run_pair[2], type = simMeasure)
  gapPenalty <- getGapPenalty(s, gapQuantile, type = simMeasure)
  Alignobj <- getAffineAlignObj(s, go = gapPenalty*goFactor, ge = gapPenalty*geFactor)
  AlignedIndices <- getAlignment(Alignobj)
  tA <- StrepChroms[[run_pair[1]]][[peptide]][[1]][["time"]]
  tB <- StrepChroms[[run_pair[2]]][[peptide]][[1]][["time"]]
  tA.aligned <- mapIdxToTime(tA, AlignedIndices[[1]][,"indexA_aligned"])
  tB.aligned <- mapIdxToTime(tB, AlignedIndices[[1]][,"indexB_aligned"])
  predictTime <- tB.aligned[which.min(abs(tA.aligned - StrepAnnot[peptide, run_pair[1]]))]
  deltaT <- predictTime - StrepAnnot[peptide, run_pair[2]]
  Err[peptide, 1] <- deltaT
}
```


```{r plotLocal}
plotErrorCurve(abs(Err), "red", xlab = "Retention time difference (in sec)", ylab = "Cumulative fraction of peptides")
```

# Plot similarity matrix and alignment path
```{r}
library(lattice)
levelplot(s, axes = TRUE, xlab = "run1 index", ylab = "run2 index")
Path <- getAlignmentPath(AlignedIndices[[1]], s)
levelplot(s, axes = TRUE, xlab = "run1 index", ylab = "run2 index", main = paste0("Alignment path through the ", simMeasure, " similarity matrix\n for ", peptide)) + latticeExtra::as.layer(levelplot(Path, col.regions = c("transparent", "green"), alpha = 1, axes = FALSE))
```



# Plot chromatograms
```{r}
library(ggplot2)
library(reshape2)
plotChromatogram <- function(data, run, peptide, StrepAnnot, printTitle =TRUE){
  df <- do.call("cbind", data[[run]][[peptide]])
  df <- df[,!duplicated(colnames(df))]
  df <- melt(df, id.vars="time", value.name = "Intensity")
  g <- ggplot(df, aes(time, Intensity, col=variable)) + geom_line(show.legend = FALSE) + theme_bw()
  if(printTitle) g <- g + ggtitle(paste0(run, ", ",peptide)) + theme(plot.title = element_text(hjust = 0.5))
  g <- g + geom_vline(xintercept=StrepAnnot[peptide, run], lty="dotted", size = 0.4)
  return(g)
}
```


Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
